\textbf{Urban mobility and socioeconomic deprivation in Latin America after COVID-19}

### 2024-02-01 manuscript/manuscript.tex
manuscript-prophet

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Socioeconomic deprivation and urban mobility in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Socioeconomic deprivation and urban mobility in Latin
America after COVID-19}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis. Our research highlights the importance of timely mobility data
with high spatiotemporal resolution to understand the long-term effects
of the pandemic and to inform equitable policy responses that address
societal challenges in urban areas.
\end{abstract}

\section{Introduction}\label{sec-intro}

@@

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL), developed by the European Comission's Joint Research Centre
(Schiavina M. 2019) to define urban areas; and the Global Gridded
Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
and Applications Centre (Columbia University 2022) from sociodemographic
and satellite data inputs. Building on existing evidence (e.g. (Rowe et
al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
has recovered returning to pre-pandemic levels of movement as
nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

@@

\section{Results}\label{sec-results}

The evolution of the levels of movement is measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units at various levels according to GADM, the
Database of Global Administrative Areas (GADM 2024). The analysis
focuses on administrative units that are within the boundaries of
functional urban areas as specified by the Global Human Settlement Layer
(GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the levels of movement are made using
the Prophet forecasting procedure. Further details are provided in
Section~\ref{sec-methods}.

\begin{figure}[H]

@@

\end{figure}%

\subsection{The heterogeneous impact of COVID-19 on urban
mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}

We analyse the evolution of the relative intensity of urban mobility
with respect to a baseline period prior to the pandemic. Specifically,
we focus the analysis on short-distance movements in urban areas to
represent local and routine mobility (Owen and Green 1992), so only
movements covering a distance of at most 70 km are considered. For a
movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

\begin{figure}[H]

@@

\end{figure}%

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the relative number of movements with respect to the
baseline. Following the initial drop in mobility, all four countries
evolve towards the recovery of baseline levels, with a generally
increasing trend. There are, however, fluctuations from the general
trend which manifest differently for each country. These fluctuations
mirror each other in the case of Argentina and Colombia, where urban
mobility sharply bounces back closer to pre-pandemic levels around July
of 2020. Chile and Mexico display more progressive patterns of recovery,
although Chile never reaches baseline levels. These fluctuations are
unique to each country and can be attributed to local factors such as
the effects of seasonality or the different stringency measures imposed
by the national governments during the pandemic.

@@

\subsection{The most deprived areas experienced the smallest drop in
mobility relative to pre-pandemic
levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the number of movements relative to the pre-pandemic baseline period and
the average RDI, at the administrative unit level. We assume that this
relationship is linear and we use a linear regression to estimate the
slope and intercept characterising the line of best fit. This is shown
for April 2020 and March 2022 in the right-hand side panels of Figure 3.
After obtaining the slope and intercept for every month, we are able to
plot the evolution of these parameters for both the observed and
forecasted data, as displayed on the left-hand-side panels of the same
Figure.

\begin{figure}[H]

@@

\caption{Evolution of the parameters characterising the relationship
between relative deprivation index and number of movements relative to
the pre-pandemic baseline period.}

\end{figure}%

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same over time despite the temporary fluctuations. The
slope captures the extent of differences in the level of urban mobility
across administrative units with varying levels of socioeconomic
deprivation. It can therefore be regarded as a measure of inequality in
mobility patterns across socioeconomic groups. A slope equal to zero
would mean that all administrative units display the same intensity of
movement regardless of their average RDI. Given the patterns observed in
Argentina, Colombia and Mexico, we find that at the beginning of the
pandemic there were notable inequalities between socioeconomic groups in
terms of the levels of urban mobility. While it has taken more than two
years for Argentina and Mexico to close the gap (their slope is close to
zero from spring 2022), inequalities persist in Chile and Colombia as of
March 2023.

@@

\section{Discussion}\label{sec-discussion}

@@

\section{Methods}\label{sec-methods}

\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}

@@

\subsection{Time series analysis}\label{time-series-analysis}

For the countries included in the analysis, the movement data is
available until March 2022. In order to gain a better understanding of
the recovery trends after the pandemic, we generate a 12-month forecast
up to March 2023. This is done using Prophet, a procedure to forecast
time series data based on an additive model where non-linear trends can
be fit with seasonality effects, holiday effects and other external
factors such as the stringency index. For the analysis, only seasonality
effects are included as they yield the most realistic predictions.

In Figures 2 and 3, outliers are removed. Outliers are defined as
administrative units which, at any given month, have a z-score greater
than 4 for the relative intensity of movement. Furthermore, a
Savitzky-Golay filter is applied to smooth the time series data, using a
length of 4 units for the filter window and order 2 polynomials to fit
the samples.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-LaurenAlexander15}
Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
{``Origin--Destination Trips by Purpose and Time of Day Inferred from
Mobile Phone Data.''} \emph{Transportation Research Part C: Emerging
Technologies} 58: 240--50.
https://doi.org/\url{https://doi.org/10.1016/j.trc.2015.02.018}.

\bibitem[\citeproctext]{ref-Cabrera-Arnau2023}
Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
Soong Moon Kang. 2023. {``Inferring Urban Polycentricity from the
Variability in Human Mobility Patterns.''} \emph{Scientific Reports} 13
(1): 5751. \url{https://doi.org/10.1038/s41598-023-33003-7}.

\bibitem[\citeproctext]{ref-RDI2022}
Columbia University, Center for International Earth Science Information
Network. CIESIN. 2022. {``Global Gridded Relative Deprivation Index
(GRDI), Version 1.''} Palisades, New York: NASA Socioeconomic Data;
Applications Center (SEDAC). \url{https://doi.org/10.7927/3xxe-ap97}.

\bibitem[\citeproctext]{ref-ECJRC22}
Commission, European, Joint Research Centre, C Bosco, S
Grubanov-Boskovic, S Iacus, U Minora, F Sermi, and S Spyratos. 2022.
\emph{Data Innovation in Demography, Migration and Human Mobility}.
Publications Office of the European Union.
\url{https://doi.org/doi/10.2760/027157}.

\bibitem[\citeproctext]{ref-UNSD}
Division, United Nations Statistical. 2019. {``Handbook on the Use of
Mobile Phone Data for Official Statistics.''}

\bibitem[\citeproctext]{ref-GADM24}
GADM. 2024. {``GADM Maps and Data.''}

\bibitem[\citeproctext]{ref-Maas19}
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\bibitem[\citeproctext]{ref-MengChuishi17}
Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
Purpose Inference with GPS Trajectories, POIs, and Geo-Tagged Social
Media Data.''} In \emph{2017 IEEE International Conference on Big Data
(Big Data)}, 1319--24.
\url{https://doi.org/10.1109/BigData.2017.8258062}.

\bibitem[\citeproctext]{ref-NiluferSariAslam21}
Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
{``ActivityNET: Neural Networks to Predict Public Transport Trip
Purposes from Individual Smart Card Data and POIs.''} \emph{Geo-Spatial
Information Science} 24 (4): 711--21.
\url{https://doi.org/10.1080/10095020.2021.1985943}.

\bibitem[\citeproctext]{ref-owen1992migration}
Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
\emph{Migration Processes and Patterns: Research Progress and
Prospects}, edited by T. Champion and T. Fielding. Belhaven Press.

\bibitem[\citeproctext]{ref-RoweBigData23}
Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
of Human Geography}, 42--47. Cheltenham, UK: Edward Elgar Publishing.
\url{https://doi.org/10.4337/9781800883499.ch09}.

\bibitem[\citeproctext]{ref-roweCalafiore2022}
Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
Understanding Human Mobility in Britain During the COVID-19 Pandemic
Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.

\bibitem[\citeproctext]{ref-GHSL19}
Schiavina M., Maffenini L., Moreno-Monroy A. 2019. {``GHS-FUA R2019A -
GHS Functional Urban Areas, Derived from GHS-UCDB R2019A, (2015),
R2019A.''} European Commission, Joint Research Centre (JRC).

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility and socioeconomic deprivation in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Urban mobility and socioeconomic deprivation in Latin
America after COVID-19}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis. Our research highlights the importance of timely mobility data
with high spatiotemporal resolution to understand the long-term effects
of the pandemic and to inform equitable policy responses that address
societal challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

@@

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL), developed by the European Comission's Joint Research Centre
(Schiavina M. 2019) to define urban areas; and the Global Gridded
Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
and Applications Centre (Columbia University 2022) from sociodemographic
and satellite data inputs. Building on existing evidence (e.g. (Rowe et
al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
has recovered returning to the pre-pandemic baseline level of movement
as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

@@

\hypertarget{sec-results}{%
\section{Results}\label{sec-results}}

The evolution of the percentage change in the number of movements is
measured with respect to a baseline period prior to the pandemic as
described in Section~\ref{sec-methods}. For the purposes of the
analysis, we aggregate the raw movement data temporally into months and
spatially into administrative units at various levels according to GADM,
the Database of Global Administrative Areas (GADM 2024). The analysis
focuses on administrative units that are within the boundaries of
functional urban areas as specified by the Global Human Settlement Layer
(GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number
of movements are made using the Prophet forecasting procedure. Further
details are provided in Section~\ref{sec-methods}.

\begin{figure}

@@

\end{figure}

\hypertarget{the-heterogeneous-impact-of-covid-19-on-urban-mobility}{%
\subsection{The heterogeneous impact of COVID-19 on urban
mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}}

We analyse the evolution of the percentage change in the number of
movements with respect to a baseline period prior to the pandemic.
Specifically, we focus the analysis on short-distance movements in urban
areas to represent local and routine mobility (Owen and Green 1992), so
only movements covering a distance of at most 70 km are considered. For
a movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

\begin{figure}

@@

\end{figure}

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the percentage change in the number of movements
with respect to the baseline. Following the initial drop in mobility,
all four countries evolve towards the recovery of baseline levels of
mobility, with a generally increasing trend. There are, however,
fluctuations from the general trend which manifest differently for each
country. These fluctuations mirror each other in the case of Argentina
and Colombia, where urban mobility sharply bounces back closer to
pre-pandemic levels around July of 2020. Chile and Mexico display more
progressive patterns of recovery, although Chile never reaches baseline
levels. These fluctuations are unique to each country and can be
attributed to local factors such as the effects of seasonality or the
different stringency measures imposed by the national governments during
the pandemic.

@@

\hypertarget{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}{%
\subsection{The most deprived areas experienced the smallest drop in
mobility relative to pre-pandemic
levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}}

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the percentage change in the number of movements relative to the
pre-pandemic baseline period and the average RDI, at the administrative
unit level. We assume that this relationship is linear and we use a
linear regression to estimate the slope and intercept characterising the
line of best fit. This is shown for April 2020 and March 2022 in the
right-hand side panels of Figure 3. After obtaining the slope and
intercept for every month, we are able to plot the evolution of these
parameters for both the observed and forecasted data, as displayed on
the left-hand-side panels of the same Figure.

\begin{figure}

@@

\caption{Evolution of the parameters characterising the relationship
between relative deprivation index and the percentage change in the
number of movements relative to the pre-pandemic baseline level of
mobility.}

\end{figure}

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same over time despite the temporary fluctuations. The
slope captures the extent of differences in the level of urban mobility
across administrative units with varying levels of socioeconomic
deprivation. It can therefore be regarded as a measure of inequality in
mobility patterns across socioeconomic groups. A slope equal to zero
would mean that all administrative units display the same level of
mobility regardless of their average RDI. Given the patterns observed in
Argentina, Colombia and Mexico, we find that at the beginning of the
pandemic there were notable inequalities between socioeconomic groups in
terms of the levels of urban mobility. While it has taken more than two
years for Argentina and Mexico to close the gap (their slope is close to
zero from spring 2022), inequalities persist in Chile and Colombia as of
March 2023.

@@

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

@@

\hypertarget{sec-methods}{%
\section{Methods}\label{sec-methods}}

\hypertarget{meta-facebook-movement-data}{%
\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}}

@@

\hypertarget{time-series-analysis}{%
\subsection{Time series analysis}\label{time-series-analysis}}

For the countries included in the analysis, the movement data is
available until March 2022. In order to gain a better understanding of
the recovery trends after the pandemic, we generate a 12-month forecast
up to March 2023. This is done using Prophet (Taylor and Letham 2017), a
procedure developed by Meta's Core Data Science team to forecast time
series data based on an additive model. Prophet stands out from other
forecasting methods due to its ability to capture recurring seasonal
trends inherent in mobility data, such as fluctuations during holidays,
or changes in the general trend due to the impact of particular
circumstances, such as the COVID-19 pandemic. It is particularly
intuitive to use compared to other models due to its automated
trend-detection capabilities, making it accessible to users with no
expertise on the underlying model, while being robust to missing data
and outliers. For the analysis in this paper, yearly seasonality effects
are included as they yield the most realistic predictions.

Figures 2 and 3 have been created by removing statistical outliers. We
define these as the administrative units which, at any given month, have
a z-score greater than 4 for the percentage change in the number of
movements. Furthermore, a Savitzky-Golay filter is applied to smooth the
time series data, using a length of 4 units for the filter window and
order 2 polynomials to fit the samples.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-LaurenAlexander15}{}}%
Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
{``Origin--Destination Trips by Purpose and Time of Day Inferred from
Mobile Phone Data.''} \emph{Transportation Research Part C: Emerging
Technologies} 58: 240--50.
https://doi.org/\url{https://doi.org/10.1016/j.trc.2015.02.018}.

\leavevmode\vadjust pre{\hypertarget{ref-Cabrera-Arnau2023}{}}%
Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
Soong Moon Kang. 2023. {``Inferring Urban Polycentricity from the
Variability in Human Mobility Patterns.''} \emph{Scientific Reports} 13
(1): 5751. \url{https://doi.org/10.1038/s41598-023-33003-7}.

\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
Columbia University, Center for International Earth Science Information
Network. CIESIN. 2022. {``Global Gridded Relative Deprivation Index
(GRDI), Version 1.''} Palisades, New York: NASA Socioeconomic Data;
Applications Center (SEDAC). \url{https://doi.org/10.7927/3xxe-ap97}.

\leavevmode\vadjust pre{\hypertarget{ref-ECJRC22}{}}%
Commission, European, Joint Research Centre, C Bosco, S
Grubanov-Boskovic, S Iacus, U Minora, F Sermi, and S Spyratos. 2022.
\emph{Data Innovation in Demography, Migration and Human Mobility}.
Publications Office of the European Union.
\url{https://doi.org/doi/10.2760/027157}.

\leavevmode\vadjust pre{\hypertarget{ref-UNSD}{}}%
Division, United Nations Statistical. 2019. {``Handbook on the Use of
Mobile Phone Data for Official Statistics.''}

\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
GADM. 2024. {``GADM Maps and Data.''}

\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
Purpose Inference with GPS Trajectories, POIs, and Geo-Tagged Social
Media Data.''} In \emph{2017 IEEE International Conference on Big Data
(Big Data)}, 1319--24.
\url{https://doi.org/10.1109/BigData.2017.8258062}.

\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
{``ActivityNET: Neural Networks to Predict Public Transport Trip
Purposes from Individual Smart Card Data and POIs.''} \emph{Geo-Spatial
Information Science} 24 (4): 711--21.
\url{https://doi.org/10.1080/10095020.2021.1985943}.

\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
\emph{Migration Processes and Patterns: Research Progress and
Prospects}, edited by T. Champion and T. Fielding. Belhaven Press.

\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
of Human Geography}, 42--47. Cheltenham, UK: Edward Elgar Publishing.
\url{https://doi.org/10.4337/9781800883499.ch09}.

\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
Understanding Human Mobility in Britain During the COVID-19 Pandemic
Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.
